---
title: "In-Class Exercise 3"
subtitle: "Spatial Analysis and Point Methods"
author: "Ooi Wen Xian"
date: "September 2, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
  cache: true
---
Edit 8/9: Fixed navigation issues

```{r}
pacman::p_load(arrow,lubridate,tidyverse,tmap,sf)
```
# 1.0 Installing Maptools

maptools is retired and binary is removed from CRAN. We can download it from Posit Public Package Manager

```
install.packages("maptools", repos = "https://packagemanager.posit.co/cran/2023-10-13")
```
# 2.0 Importing Geospatial Data

In this section, st_read() of sf package will be used to import the three geospatial data sets mentioned in previous section into R environment.
```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson")
```
```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```
# 3.0 Creating Coastal Outline

```{r}
sg_sf <- mpsz_sf %>%
  st_union()

plot(sg_sf)
```
# 4.0 Importing acled data

Myammar has 2 different coordinate reference systems, hence we would be using one of them for this example.

```{r}
acled_sf <- read_csv("data/ACLED_Myanmar.csv") %>%
  st_as_sf(coords=c("longitude", "latitude"),
           crs=4326) %>%
  st_transform(crs=32647) %>%
  mutate(event_date = dmy(event_date))
```
```{r}
tmap_mode('plot')  
# Used plot mode for faster rendering, as file took too long to render with interactive map
acled_sf %>% 
  filter(year == 2023 | event_type == 'Political violence') %>% 
  tm_shape() + tm_dots()
tmap_mode('plot')
```