---
title: "In-Class Exercise 2"
author: "Ooi Wen Xian"
date: "August 26, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---

# 1.0 Importing geospatial data

Data is extracted from this source:

Master Plan 2014 Subzone Boundary (Web) (i.e. MP14_SUBZONE_WEB_PL) in ESRI shapefile format. It can be downloaded at data.gov.sg. This is a geospatial data. It consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2014.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 1.1 Importing the shapefile

```{r}
mpsz_shp <- st_read(dsn = "C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data", layer = "MP14_SUBZONE_WEB_PL")
```
Checking class of mpsz_shp
```{r}
class(mpsz_shp)
```
## 1.2 Importing the kml file

```{r}
mpsz_kml <- st_write(mpsz_shp, "C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/MP14_SUBZONE_WEB_PL.kml", delete_dsn = TRUE)
```
```{r}
mpsz_kml
```

```{r}
mpsz2019_kml <- st_read("C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/mpsz-2019/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```
```{r}
mpsz2019_shp <- st_read(dsn = "C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/mpsz-2019", layer="mpsz-2019") %>%
  st_transform(crs=3414)

```
```{r}
mpsz2019_shp <- st_read(dsn = "C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/mpsz-2019", layer="mpsz-2019") %>%
  st_transform(crs=3414)

```

```{r}
preschool <- st_read(dsn = "C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/PreSchoolsLocation.kml") 
```

```{r}
st_crs(mpsz2019_shp)
```

```{r}
popdata <- read.csv("C:/wamp64/www/crediblues/IS415-GAA/In-Class_Ex/In-Class_Ex02/data/respopagesextod2023.csv")
```

```{r}
popdata2023 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP)
colnames(popdata2023)
```

