---
title: "In-Class Exercise 4"
author: "Ooi Wen Xian"
date: "September 9, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
---
# 1.0 Installing and Loading the R packages
```{r}
pacman::p_load(sf, raster, spatstat, tmap, tidyverse, sparr)
```

# 2.0 Importing and preparing study area
```{r}
kbb <- st_read(dsn = "data/rawdata", layer="Kepulauan_Bangka_Belitung")
```
```{r}
#| eval: false
view(kbb)
```

There are 3 issues with kbb. 
1. Boundaries should be grouped up
2. Polygon Z value may cause issues later on
3. Data is in wgs84 and not in a projected crs

```{r}
kbb_sf <- st_read(dsn = "data/rawdata", layer="Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs=32748)
```
Converting to OWIN.
as.owin() is used to convert kbb to an owin object.

```{r}
kbb_owin <-as.owin(kbb_sf)
kbb_owin
```
```{r}
class(kbb_owin)
```
# 3.0 Importing and Preparing FOrest Fire Data

```{r}
fire_sf <- read.csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords=c("longitude", "latitude"), crs=4326) %>%
  st_transform(crs=32748)
```

```{r}
fire_sf$acq_date <- as.Date(fire_sf$acq_date, format = "%Y/%m/%d")
```

```{r}
fire_sf <- fire_sf %>%
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, label=TRUE, abbr=FALSE))
  
```

```{r}
# polygon should be plotted first before the points
# polygon, line, points
tmap_mode("plot")
tm_shape(kbb_sf)+tm_polygons()+tm_shape(fire_sf)+tm_dots()
```
```{r}
tmap_mode("plot")
tm_shape(kbb_sf)+tm_polygons()+tm_shape(fire_sf)+tm_dots(size = 0.1)+
  tm_facets(by="Month_fac", free.coords = FALSE, drop.units = TRUE)
# if you do not have free.coords = FALSE, drop.units = TRUE, the maps would zoom to where you have the data.
# month_fac is important as it ensures the data is sorted monthly and not alphabetically
```
# 4.0 Computing STKDE by month

```{r}
fire_month <- fire_sf %>% select(Month_num)
```

sparr is not just looking at events in isolation, but it allows you to look at the contrast between events and control

```{r}
fire_month_ppp<- as.ppp(fire_month)
fire_month_ppp
```
```{r}
summary(fire_month_ppp)
```
```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```
```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```
lamda --> saying that there is 0.0304 months

```{r}
#| fig.width: 12
#| fig.height: 10
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims)(
  plot(st_kde, i,
       override.par=FALSE,
       fix.range=TRUE,
       main=paste("KDE at month", i))
)
```








